# Randomization

```{r, include=FALSE, echo = FALSE,message = FALSE, error = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# <!-- ---------------------------------------------------------------------- -->
# <!--                    1. load the required packages                       -->
# <!-- ---------------------------------------------------------------------- --> 

## if(!require(psych)){install.packages("psych")}

packages<-c("tidyverse", "kableExtra", "gtsummary","inTextSummaryTable",
            "Hmisc","htmltools","clinUtils")
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
ipak(packages)
 
# <!-- ---------------------------------------------------------------------- -->
# <!--                        2. Basic system settings                        -->
# <!-- ---------------------------------------------------------------------- -->
# setwd(dirname(rstudioapi::getSourceEditorContext()$path))
```


<!-- ----------------------------------------- -->
<!-- -----------Mind Map of ToC--------------- -->
<!-- ----------------------------------------- -->


```{r,echo = F,message = FALSE, error = FALSE, warning = FALSE}
library(mindr)
input <- rstudioapi::getSourceEditorContext()$path 
input_txt <- readLines(input, encoding = "UTF-8")
## Widget Output
mm_output <- mm(input_txt, 
                output_type = c("widget"),
                root = "")
mm_output$widget
```




### Simple randomization

Randomization based on a single sequence of random assignments is known
as simple randomization.

> 对于两个治疗组（对照组与治疗组），硬币的侧面（即正面-对照组，反面-治疗）确定每个受试者的分配。其他方法包括使用经过洗牌的纸牌（例如，偶数控制，奇数处理）或掷骰子（例如，低于并等于3控制，超过3处理）这种随机方法简单易行，可在临床研究中实施。在大型临床研究中，可以相信简单随机会在组之间生成相似数量的主题。但是，在相对较小的样本量临床研究中，随机化结果可能会出现问题，从而导致组中参与者的数量不相等。

```{r,echo = T,message = FALSE, error = FALSE, warning = FALSE}
set.seed(888)
treatment <- c("A","B")
simple.list <- sample(treatment, 20, replace=TRUE)
cat(simple.list,sep="\n")
table(simple.list)
```

### Block randomization

The block randomization method is designed to randomize subjects into
groups that result in equal sample sizes. This method is used to ensure
a balance in sample size across groups over time. Blocks are small and
balanced with predetermined group assignments, which keeps the numbers
of subjects in each group similar at all times. The block size is
determined by the researcher and should be a multiple of the number of
groups (i.e., with two treatment groups, block size of either 4, 6, or
8). Blocks are best used in smaller increments as researchers can more
easily control balance.

> 块随机化方法旨在将受试者随机分组，从而产生相同的样本量。使用此方法可确保随着时间的推移，各个组之间的样本量保持平衡。区块很小，并且通过预定的小组分配来平衡，这使每个小组中的被试人数始终保持相似。区块大小由研究人员确定，并且应为组数的倍数（即，对于两个治疗组，区块大小为4、6或8）。最好以较小的增量使用块，因为研究人员可以更轻松地控制平衡。

After block size has been determined, all possible balanced combinations
of assignment within the block (i.e., equal number for all groups within
the block) must be calculated. Blocks are then randomly chosen to
determine the patients' assignment into the groups.

> 确定块大小后，必须计算该块内分配的所有可能平衡组合（即，该块内所有组的数量均相等）。然后随机选择块来确定患者的分组。

**Example 以区组长度4为例**

-   一个区组内的4个研究对象可以有6种排列方式：1. AABB, 2. ABAB, 3.
    ABBA, 4. BAAB, 5. BABA, 6. BBAA
-   确定好所有的排列形式后，接下来需要将6个区组随机排列。我们可以用各种方式（如SPSS、Excel、SAS等）产生一串随机数字.
    因为只有6种排列方式，因此可以只选择1-6之间的数字，25126423121362555343526422
-   按照上述随机数字排列区组

**BLOCKED RANDOMIZATION USING THE PROC PLAN**

    * randomization within block;
    %macro ran_block (blockNum);
    …
    %if &seed= %then %do;
     proc plan;
     factors N=&blocksize / noprint;
     output data=One out=Two
     run;
    %end;
    %else %do;
     Proc plan seed=%eval (&seed+blockNum);
     Factors N=&blockSize / noprint;
     Output data=One out=Two;
     Run;
    %end;
    …
    %mend ran_block;

**BLOCKED RANDOMIZATION USING psych Package**

```{r,echo = T,message = FALSE, error = FALSE, warning = FALSE}
library("psych")
my.cond <- block.random (n=96,c( drug=2,time=3,imp=2)) 
headtail(my.cond) 
## Visualizing block randomization
pairs.panels (my.cond)
```

**BLOCKED RANDOMIZATION USING blockrand Package**

```{r,echo = T,message = FALSE, error = FALSE, warning = FALSE}
### Generate a block randomization for a clinical trial;
library("blockrand")
set.seed(888)
block.list <- blockrand(n=20, num.levels = 2,block.sizes = c(2,2))
block.list
block.list2 <- blockrand(n=20, num.levels = 2,block.sizes = c(1,2))
block.list2
plot(block.list2)
```

### Stratified randomization

The stratified randomization method addresses the need to control and
balance the influence of covariates. This method can be used to achieve
balance among groups in terms of subjects' baseline characteristics
(covariates). Specific covariates must be identified by the researcher
who understands the potential influence each covariate has on the
dependent variable. Stratified randomization is achieved by generating a
separate block for each combination of covariates, and subjects are
assigned to the appropriate block of covariates. After all subjects have
been identified and assigned into blocks, simple randomization is
performed within each block to assign subjects to one of the groups.

> 分层随机方法解决了控制和平衡协变量影响的需求。此方法可用于在受试者的基线特征（协变量）方面实现组间的平衡。研究人员必须确定特定的协变量，他们必须了解每个协变量对因变量的潜在影响。分层随机化是通过为协变量的每种组合生成一个单独的块来实现的，并将主题分配给合适的协变量块。在确定了所有主题并将其分配到块中之后，在每个块内执行简单的随机化以将主题分配给其中一个组。

$$
\begin{array}{|l|c|c|}
\hline \text { Stratum } & \text { Age Group } & \text { Severity Score } \\
\hline 1 & \geq 50 \text { years } & \geq 7 \\
\hline 2 & <50 \text { years } & \geq 7 \\
\hline 3 & \geq 50 \text { years } & <7 \\
\hline 4 & <50 \text { years } & <7 \\
\hline
\end{array}
$$

```{r,echo = T,message = FALSE, error = FALSE, warning = FALSE}
over50.severe.list <- blockrand(n=100, num.levels = 2,
                                block.sizes = c(1,2,3,4), 
                                stratum='Over 50, Severe',id.prefix='O50_S', block.prefix='O50_S')
headtail(over50.severe.list)
```
